---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gentleman <img src="man/figures/logo.svg" style="float:right; height:200px;" />

<!-- badges: start -->
<!-- badges: end -->

The **gentleman** package provides a suite of modules to help with data preparation, descriptive statistics, modeling, and publication formatting. The Descriptives module is extensible to accommodate custom functions to generate tables and p-values (or other useful statistic to compare groups). The Models module can be used to generate lavaan model strings for mediation and cross-lagged models.

## Installation

You can install gentleman from the public repository on GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("patc3-phd/gentleman")
```

## Using gentleman

Once installed, you can use the package by loading it into your R session.

```{r, eval=FALSE}
library(gentleman)
```

## Modules

The package is composed of several modules that accomplish related tasks (see the [reference page](https://d22consulting.com/gentleman/reference/index.html) for a complete list of functions):

* Data preparation
* Descriptive statistics
* Models
* Clusters
* Machine learning
* Logger
* Helpers



### 1. Data Preparation

With the data preparation module, you can clean and manipulate data.frames, and recode variables and modify factors. For example:

```{r, eval=FALSE}
# Remove non-ASCII characters and variables with too many missing values,
# cast character vars as factors, and remove factors with too many categories
df <- df |> 
  remove_non_ascii_from_df() |> 
  remove_vars_with_too_many_missing(pmissing=.80) |> 
  cast("character", "factors") |> 
  remove_blank_factor_levels() |> 
  remove_factors_with_too_many_levels(maxlevels=20)

# Recode 'Nationality' using a lookup table in an Excel file
df$Nationality <- df$Nationality |> 
  recode_using_excel_map("map.xlsx", sheet="Nationality")

# Use effect coding for factors in regression and other models
options(contrasts=c(unordered="contr.dummy_or_effect", 
                    ordered="contr.poly"))
```



### 2. Descriptive Statistics

The package is equipped with an extensive and customizable module for descriptive statistics. In particular, you can generate **publication-ready tables** of descriptive statistics for numeric and categorical variables, optionally with added columns for group-specific statistics and group comparison.

For example:

```{r, eval=FALSE}
# Get a table of descriptive statistics for numeric variables
# and test for group differences by treatment group
df |> get_desc_table(
  vars=c("age", "score"),
  tbl_fn=tbl_fn_num,
  group="TreatmentGroup",
  ana_fn=ana_fn_aov
)

# Get a table of descriptive statistics for factor variables
# and test for group differences by treatment group
df |> get_desc_table(
  vars=c("Nationality", "Diagnostic"),
  tbl_fn=tbl_fn_fac,
  group="TreatmentGroup",
  ana_fn=ana_fn_chisq
)
```

The `get_desc_table()` function is the workhorse of the module, which takes as arguments:

* **`tbl_fn`**: A function that returns a table of statistics for given variables; the package comes with `tbl_fn_num()` for numeric variables, and `tbl_fn_fac()` for categorical (factor) variables
* **`ana_fn`**: A function that returns a table of test statistics (typically *p*-values) to compare groups on a set of variables; the package comes with `ana_fn_aov()` and `ana_fn_rm_aov()` for (repeated) numeric variables (*p*-values from ANOVA), and `ana_fn_chisq()` for categorical (factor) variables (*p*-values from Chi-square tests)

You can develop custom `tbl_fn` and `ana_fn` functions as you wish and pass them as arguments to `get_desc_table()`, which will take care of the rest.

The module can also be used to detect which variables vary within persons or by groups, and to plot densities by groups.

```{r, eval=FALSE}
# Test mean differences between subscales
df |> compare_pairs_of_vars(c("x1", "x2", "x3"))

# Get all variables that significantly vary by groups
df |> get_sig_differences_between_groups(group="TreatmentGroup")

# Plot densities by group
df |> plot_density_by_groups(group="TreatmentGroup")
```



### 3. Models

The models module can generate `lavaan` syntax to estimate complex mediation and cross-lagged models:

```{r, eval=FALSE}
library(lavaan)

# Mediation model with one x, one mediator, and one y
get_mediation_model("x1", "x2", "x3") |>
  sem(df) |>
  summary()

# Cross-lagged model with 3 different variables assessed at 3 occasions each
vars <- list(
  c("x1", "x2", "x3"),
  c("y1", "y2", "y3"),
  c("z1", "z2", "z3")
)
get_crosslagged_model(vars, random_intercepts=FALSE) |>
  sem(df) |>
  summary()
```

The module can also be used to decompose 2-way interactions in different types of models (like `lm` and `lmer`):

```{r, eval=FALSE}
fit <- lm(SelfEsteem ~ Gender*AttitudeTowardsSchool, df)
fit |> decompose_interaction(x1="Gender", x2="AttitudeTowardsSchool")
```

Similar to the descriptives module, you can generate **publication-ready tables** for different types of models that you can copy-paste, for example into Excel. Effects are formatted with 2 decimals and no leading zero, and significance is indicated with common standards using asterisks. Currently, publication tables are available for `lavaan` models and models that can be summarized using `broom` (like `lm` and `glm`).

```{r, eval=FALSE}
# Generate a publication table for two regressions using lavaan 
# and full-information maximum likelihood estimation
library(lavaan)
fit_lavaan <- list()
fit_lavaan$y1 <- sem("y1 ~ x1+x2", df, missing="ml.x")
fit_lavaan$y2 <- sem("y2 ~ x1+x2", df, missing="ml.x")
pub_lavaan <- fit_lavaan |> make_pub_table_from_lavaan_models()

# Generate a publication table for two regressions with OLS estimation 
# and listwise deletion of missing data
library(broom)
fit_lm <- list()
fit_lm$y1 <- lm(y1 ~ x1+x2, df) |> tidy()
fit_lm$y2 <- lm(y2 ~ x1+x2, df) |> tidy()
pub_lm <- fit_lm |> make_pub_table_from_broom_tidy()
```

Once you've generated publication tables from several models, you can compare significant effects between publication tables:

```{r, eval=FALSE}
# Sensitivity analysis: effect of missing data
pub_lm |> compare_sig_effects_in_two_pub_tables(pub_lavaan)
```




### 4. Clusters

Clustering can be performed easily with mixed types of variables and missing data. The module uses the **cluster** package to first calculate distance matrices based on Gower distances, then generate clusters around medoids. The desired number of clusters can be specified, or it can be determined automatically via the **NbClust** package.

```{r, eval=FALSE}
# Generate 3 clusters using all variables
df |> get_cluster(k=3)
```

The module also implements an iterative variable selection procedure, where variables are kept in the clustering process only if they are related to the final cluster assignments.

```{r, eval=FALSE}
# Generate clusters with variable selection using an optimal number of clusters
# and plot densities of variables on which clusters significantly vary
df |>
  add_cluster_assignment(k=NULL, 
                         maxit=10, 
                         return_df_cluster_instead=TRUE) |>
  plot_density_by_groups(group="Cluster")
```




### 5. Machine Learning

The machine learning module provides functions to support the development of ML pipelines, particularly for supervised learning. You can split into training and test sets, scale numeric features on both sets using the information from the training set, and remove the target from the test set to prevent target leakage (an implementation of the method discussed [here](https://en.d22consulting.com/quantcafe/preventing-target-leakage)).

```{r, eval=FALSE}
# Split into train-test sets, scale numeric features,
# and remove target from test set to prevent target leakage
tt <- df |> 
  ttsplit(prop_train=.67) |> 
  scale_numeric_features_in_train_and_test() |> 
  remove_target_from_test_and_add_ref_to_env(target="y1", 
                                             unique_id="id", 
                                             ref_name="target_values")
```


The module also offers an interface to generate and process an AutoML model using the **h2o** library, and can be used to add predictions to a test set and retrieve variable importances. The module also comes with an AutoML pipeline built-in, which takes in a `data.frame`, splits it into training and test sets, prepares them for h2o, and requests an AutoML model and obtains predictions and variable importances.

For example:

```{r, eval=FALSE}
# AutoML pipeline with default configuration (random forest),
# 70%-30% train-test split, scaled numeric features,
# confidence levels for predictions, and variable importances
pip <- df |> run_automl_pipeline(target="y1",
                                 prop_train=.7,
                                 scale_numeric_features=TRUE,
                                 add_confidence_level=TRUE,
                                 plot_and_print_variable_importances=TRUE)
```





### 6. Logger

The logger module offers a simple way to log console output to a text file on disk. All logs are saved in a given directory in a folder called `gentleman_logs`, and the file name can be optionally specified (by default the file name will be the current date and time). Logs can be updated by passing a logger object. The module can also be used to retrieve information about the latest commit in a git repository.

```{r, eval=FALSE}
# Simple one-time log
logtext(expr={
  print("model without covariates:")
  lm(y1 ~ x1, df) |> summary() |> print()
  
  print("model with covariates")
  lm(y1 ~ x1 + x2 + x3, df) |> summary() |> print()
})

# Create a logger with git information and update log
logger <- get_logger(fname="models with and without covariates", add_git_info=TRUE)
logger |> logtext(expr={
  print("model without covariates:")
  lm(y1 ~ x1, df) |> summary() |> print()
})
logger |> logtext(expr={
  print("model with covariates")
  lm(y1 ~ x1 + x2 + x3, df) |> summary() |> print()
})
```



### 7. Helpers

Other miscellaneous helper functions are available. Some are aliases and wrappers (like `view()`, `copy()` and `copy2()`), others are utility functions that can fasten development (for example, you can use `replace_in_nested_list()` to find and replace values in a complex list of vectors with a nested structure while preserving the structure).

See the [complete list of functions](https://d22consulting.com/gentleman/reference/index.html) for more details.

