---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gentleman <img src="man/figures/logo.svg" style="float:right; height:200px;" />

<!-- badges: start -->
<!-- badges: end -->

The *gentleman* package provides a suite of modules to help with data preparation, descriptive statistics, modeling, and publication formatting. The Descriptives module is extensible to accommodate custom functions to generate tables and p-values (or other useful statistic to compare groups). The Models module can be used to generate lavaan model strings for mediation and cross-lagged models.

## Installation

You can install the development version of gentleman from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("patc3-phd/gentleman")
```

## Using gentleman

Once installed, you can use the package by loading it into your R session.

```{r example}
library(gentleman)
```
